---
title: "EWAS script"
output: html_notebook
---

Step 1: Load packages. 
```{r}
library(data.table)
library(ggplot2)
library(future.apply)
library(milorGWAS)
library(xlsx)
library(dplyr)
library(lmerTest)
library(future.apply)
library(tidyverse)
library(EpiDISH)
library(QCEWAS)
library(ENmix)
```

Step 2: Load Beta and Meta files, and modify accordingly. 
```{r}
##Load beta and meta files. 
beta312 <- readRDS("~/data/Round 4 Linear model/Beta312")
meta312 <- read.xlsx("Metafile.xls", sheetIndex = 1)
beta312=beta312[,match(meta312$Basename,colnames(beta312))] #ensure that the order of samples is the same as in spreadsheet
CpGmanifest <- readRDS("/data/amin/Round 5 Linear model - interaction variable/CpGmanifest")

## Rename columns
meta312 = left_join(meta312, celltypePCAvaluesonly, by = "Basename")
meta312$Year <- as.character(meta312$Year)
meta312$Id <- as.character(meta312$Id)
meta312$case <- 0
meta312$case[meta312$incidentcase==1] <- 1
meta312$case[meta312$establishedcase==1] <- 1
meta312 <- rename(meta312, "fitted" = "transitioned" )

## Creating beta and meta files of incident kidney disease for model A
##Create meta file of incident cases only
metaIncidentKidneydisease <- meta312[meta312$establishedcase!=1,]
### Generate beta file corresponding to above meta file
betaIncidentKidneydisease=beta312[,colnames(beta312) %in% metaIncidentKidneydisease$Basename] #choose the subset of samples 
betaIncidentKidneydisease=betaIncidentKidneydisease[,match(metaIncidentKidneydisease$Basename,colnames(betaIncidentKidneydisease))]###make sure columns and rows in same order
```

Step 3: Generate cell-type PCAs
```{r}
data(centDHSbloodDMC.m)
Celltypefraction <- epidish(beta.m = beta312, ref.m = centDHSbloodDMC.m, method = "RPC")$estF
Celltypefraction <- as.data.frame(Celltypefraction)
celltypePCA <- prcomp(Celltypefraction, center= TRUE, scale.=TRUE)
summary(celltypePCA)
##Extract PCAs of IDs into a new dataframe
celltypePCAvaluesonly <- as.data.frame(celltypePCA[["x"]])
celltypePCAvaluesonly= mutate(celltypePCAvaluesonly, "Basename" = row.names(celltypePCAvaluesonly))
saveRDS(celltypePCAvaluesonly, file="celltypePCAvaluesonly_epidish")
```

Step 4: Cell fractions of models.  

```{r}
Celltypefraction= mutate(Celltypefraction, "Basename" = row.names(Celltypefraction))

##For Model A
Celltypefraction_modelA <- left_join(metaIncidentKidneydisease[,c("Basename","incidentcase")], Celltypefraction, "Basename") 
Celltypefraction_modelA <- select(Celltypefraction_modelA, -"Basename")
Celltypefraction_modelA$incidentcase <- as.character(Celltypefraction_modelA$incidentcase)
Celltypefraction_modelA_long <- melt(Celltypefraction_modelA, id = "incidentcase") 
ggplot(Celltypefraction_modelA_long, aes(x = variable, y = value, fill= incidentcase )) + geom_boxplot() + labs(y="Fraction", x="Cell-type", title ="Cell types model A") +
  theme(plot.title = element_text(hjust = 0.5)) + 
  guides(fill=guide_legend(title="Case")) +
scale_fill_manual(values = c("0" = "#56B4E9", "1" = "#D55E00"))
ggsave("Model A cell types.png")

##For Model B
Celltypefraction_modelB <- left_join(meta312[,c("Basename","transitioned")], Celltypefraction, "Basename") 
Celltypefraction_modelB <- select(Celltypefraction_modelB, -"Basename")
Celltypefraction_modelB$transitioned <- as.character(Celltypefraction_modelB$transitioned)
Celltypefraction_modelB_long <- melt(Celltypefraction_modelB, id = "transitioned") 
ggplot(Celltypefraction_modelB_long, aes(x = variable, y = value, fill= transitioned )) + geom_boxplot() + labs(y="Fraction", x="Cell-type", title ="Cell types model B") +
  theme(plot.title = element_text(hjust = 0.5)) + 
  guides(fill=guide_legend(title="Case"))+
scale_fill_manual(values = c("0" = "#56B4E9", "1" = "#D55E00"))
ggsave("Model B cell types.png")


```


Step 5: Create function for mixed-effect models. 
```{r}
lmer_func=function(dataset,pd,rhs_formula,testL)
{
  pd$cpg=as.numeric(t(dataset))
  pd=as.data.frame(pd)
  if (!(is.null(testL))) {
    me_model=lmer(as.formula(paste("cpg ~ ", rhs_formula,sep="")),pd) # create model object
    r1=fixef(me_model) # model coefficients for fixed effects
    r2=contest1D(model = me_model,L = testL,confint = TRUE) 
    c(as.numeric(r1),as.numeric(r2))  
  } else {
    me_model=lm(as.formula(paste("cpg ~ ", rhs_formula,sep="")),pd)
    mes = summary(me_model)
    df = me_model$df.residual
    ms = mes$coefficients
    mc = confint(me_model)
    traitInd = nrow(mc)
    c(ms[,1],ms[traitInd,c(1,2)],df,ms[traitInd,3],mc[traitInd,],ms[traitInd,4])
  }
}

ewas_mixed_model=function(dataset,meta,trait_name,fixed_effects,random_effects,fit_slope=FALSE)
{
  ###############################################
  # define fixed and random effects for the model
  ###############################################
  fixed_effects_=c(fixed_effects,trait_name) # trait_name is case-control variable name
  
  fe_formula=paste(fixed_effects_,collapse = " + ")
  if (length(random_effects)>0) {
    if (fit_slope){
      re_formula=paste("(",paste(trait_name,random_effects,sep = "|",collapse = ") + ("),")",sep="")}
    else{
      re_formula=paste("(",paste("1",random_effects,sep = "|",collapse = ") + ("),")",sep="")}
    rhs_formula = paste(fe_formula,re_formula,sep = " + ")
    testL = rep(0,length(fixed_effects_))
    testL = c(testL,1)
  } else {
    rhs_formula = fe_formula
    testL = NULL
    
  }  
  print(paste("Effects are defined. Fixed effects are ", paste(fixed_effects_,sep=', '), ". Random effects are ", paste(random_effects,sep=', '), ".", sep=""))
  
  ############################### 
  # prepare data to run the model
  ###############################  
  pd=as.data.frame(meta)[c(fixed_effects_,random_effects)] # extract dataset to from metadata to plug in into model
  ###
  pdc = colnames(pd)
  # Drop samples without metadata
  exclude_samples=(as.data.frame(pd)[trait_name]!="") # boolean vectors with FALSE for samples with missing trait
  pd=pd[exclude_samples,]
  dataset=dataset[,exclude_samples]
  pd=as.data.frame(pd)
  colnames(pd) = pdc
  dataset=as.data.frame(dataset)
  ###
  # run model
  print("planning multisession...")
  print(Sys.time())
  plan("future::multisession")
  print("multisession planned, sorry for the delay. Now running EWAS.")
  tts=future_apply(as.data.frame(dataset), MARGIN = 1, FUN = lmer_func,pd=pd, rhs_formula=rhs_formula, testL = testL)
  print("nearly done...")
  tts=t(tts)
  # "sexMale","w3_age","age_diff","w3_diabetes1"
  colnames(tts)=c("(Intercept)",fixed_effects_,"Estimate","Std. Error","df","t value","lower","upper","Pr(>|t|)")
  #print("done!")
  return(tts)
}
```

Step 6: Run models looking at incident cases, established cases and all cases. 

```{r}
fixed_effects=c("PC1","PC2","PC3","Smoke","Sex","Age")
random_effects= c("Id", "Sentrix_ID","Sentrix_Position", "Year")

trait_name="transitioned" 
# Model A : linear mixed-effect model of post transitioned cases (n=73) vs pre-transitioned incident cases (n=54) and healthy controls (n=149) model of transitioned incident cases vs healthy controls. Fixed effects: Age, Sex, Smoker status and cell types. Random effects: Year, Id, Slide & Array position.
modelA=ewas_mixed_model(dataset=as.data.frame(betaIncidentKidneydisease),meta=metaIncidentKidneydisease,trait_name = trait_name,fixed_effects = fixed_effects, random_effects = random_effects,fit_slope = FALSE)
print(Sys.time())
print("model A is done!")

# Model B : linear model of Established cases (n=36) and incident cases post-transition (n=73) vs incident cases pre-transition (n=54) and controls (n=149). Fixed effects: Age, Sex, Smoker status and first three cell-type PCA. Random effects: Year, Id, Slide & Array position.
modelB=ewas_mixed_model(dataset=as.data.frame(beta312),meta=meta312,trait_name = trait_name,fixed_effects = fixed_effects, random_effects = random_effects,fit_slope = FALSE)
print(Sys.time())
print("model B is done!")



modelA<-as.data.frame(modelA)
modelB<-as.data.frame(modelB)
fwrite(as.data.frame(modelA),paste("modelA.csv"),row.names = TRUE,col.names = TRUE)
fwrite(as.data.frame(modelB),paste("modelB.csv"),row.names = TRUE,col.names = TRUE)


```

Step 7: Functional annotation of top 20 CpGs from each model. 
```{r}
## How to identify CpGs with lowest p-vlaues - model A
lowestpvalues_modelA<-top_n(modelA , -20, )
lowestpvalues_modelA<-lowestpvalues_modelA[, c("transitioned","lower","upper","Pr(>|t|)")]
lowestpvalues_modelA <- rownames_to_column(lowestpvalues_modelA)
CpGinfomodelA <- left_join(lowestpvalues_modelA, CpGmanifest, by = c("rowname" = "IlmnID"))
CpGinfomodelA<- CpGinfomodelA[, c("rowname","transitioned","Pr(>|t|)","UCSC_RefGene_Name","UCSC_RefGene_Group", "CHR_hg38", "Start_hg38" )]
CpGinfomodelA$Base <- CpGinfomodelA$Start_hg38+1
CpGinfomodelA <- subset(CpGinfomodelA, select =-c(Start_hg38))
CpGinfomodelA<-arrange(CpGinfomodelA, (`Pr(>|t|)`))
write.csv(CpGinfomodelA, "Top 20 CpGs Model A.csv")


## How to identify CpGs with lowest p-vlaues - model B
lowestpvalues_modelB<-top_n(modelB , -20, )
lowestpvalues_modelB<-lowestpvalues_modelB[, c("transitioned","lower","upper","Pr(>|t|)")]
lowestpvalues_modelB <- rownames_to_column(lowestpvalues_modelB)
CpGinfomodelB <- left_join(lowestpvalues_modelB, CpGmanifest, by = c("rowname" = "IlmnID"))
CpGinfomodelB<- CpGinfomodelB[, c("rowname","transitioned","Pr(>|t|)","UCSC_RefGene_Name","UCSC_RefGene_Group", "CHR_hg38", "Start_hg38" )]
CpGinfomodelB$Base <- CpGinfomodelB$Start_hg38+1
CpGinfomodelB <- subset(CpGinfomodelB, select =-c(Start_hg38))
CpGinfomodelB<-arrange(CpGinfomodelB, (`Pr(>|t|)`))
write.csv(CpGinfomodelB, "Top 20 CpGs Model B.csv")



```


Step 8: Save data frames of p-values generated from mixed-effect models and generate pdf with QQ-plots of p-values. 
```{r}
pdf(file = "qqplots.pdf")
qqplot.pvalues(modelA[,ncol(modelA)],main="Mixed-effect model A", ylim=c(0,8))
qqplot.pvalues(modelB[,ncol(modelB)],main="Mixed-effect model B", ylim=c(0,8))


dev.off()

P_lambda(modelA$`Pr(>|t|)`)
P_lambda(modelB$`Pr(>|t|)`)


```

Step 9: Volcano plots.
```{r}
## Model A
png(filename = "volc model A.png", width=600, height=600)
ggplot(data=modelA, aes(x=transitioned, y=-log10(`Pr(>|t|)`))) + 
  geom_point(col="firebrick2", size=1, show.legend = FALSE) + 
  theme(
    axis.title.x = element_text(size = 27),
    axis.text.x = element_text(size = 23),
    axis.title.y = element_text(size = 27),
    axis.text.y = element_text(size = 23),
    plot.margin = margin(t= 5, r=20, b=5, l=5),
    legend.position = "none" )+
  xlim(-0.0375, 0.0375)+
  ylim(0,8)+
  labs(y= "-log(p-value)", x = "Change in methylation") +
  geom_hline(aes(yintercept=-log10(0.000000036), col="red",)) +
  geom_text(aes(0,-log10(0.000000036), label = "Threshold of signifiance", vjust = -0.5), size = 8)
dev.off()

## Model B
png(filename = "volc model B.png", width=600, height=600)
ggplot(data=modelB, aes(x=transitioned, y=-log10(`Pr(>|t|)`))) + 
  geom_point(col="firebrick2", size=1, show.legend = FALSE) + 
  theme(
    axis.title.x = element_text(size = 27),
    axis.text.x = element_text(size = 23),
    axis.title.y = element_text(size = 27),
    axis.text.y = element_text(size = 23),
    plot.margin = margin(t= 5, r=20, b=5, l=5),
    legend.position = "none" )+
  xlim(-0.0375, 0.0375)+
  ylim(0,8)+
  labs(y= "-log(p-value)", x = "Change in methylation") +
  geom_hline(aes(yintercept=-log10(0.000000036), col="red",)) +
  geom_text(aes(0,-log10(0.000000036), label = "Threshold of signifiance", vjust = -0.5), size = 8)
dev.off()


```


Step 10: ipDMR

```{r}
CpGessential <- subset(CpGmanifest, select = c(IlmnID, CHR_hg38, Start_hg38, End_hg38))
modelA<-rownames_to_column(modelA)
modelB<-rownames_to_column(modelB)

modelA_ipDMR <- left_join(modelA, CpGessential, by = c("rowname"="IlmnID"))
modelB_ipDMR <- left_join(modelB, CpGessential, by = c("rowname"="IlmnID"))

modelA_ipDMR<-dplyr::rename(modelA_ipDMR, "probe"= "rowname", "start" = "Start_hg38", "end"="End_hg38", "p"="Pr(>|t|)", "chr" = "CHR_hg38" )
modelB_ipDMR<-dplyr::rename(modelB_ipDMR, "probe"= "rowname", "start" = "Start_hg38", "end"="End_hg38", "p"="Pr(>|t|)", "chr" = "CHR_hg38" )
```

Step 11: Run the ipdmr function on each Model.
```{r}
setwd("Model A")
ipdmr(modelA_ipDMR, dist.cutoff = 1000, bin.size = 50, seed= 0.05)
setwd("Model B")
ipdmr(modelB_ipDMR, dist.cutoff = 1000, bin.size = 50, seed= 0.05)


##load the files into R
ModelA_DMR <- read_csv("Model A/resu_ipdmr.csv")
ModelB_DMR <- read_csv("Model B/resu_ipdmr.csv")

##Generate table of all relevant CpGs
##model A
modelA_DMR_CpGs<-(na.omit(modelA_ipDMR[((modelA_ipDMR$chr=="chr5" & modelA_ipDMR$start>=23507130 & modelA_ipDMR$end<=23507400) |(modelA_ipDMR$chr=="chr7" & modelA_ipDMR$start>=38464800 & modelA_ipDMR$end<=38465510)                                          |(modelA_ipDMR$chr=="chr10" & modelA_ipDMR$start>=71323360 & modelA_ipDMR$end<=71324200)|(modelA_ipDMR$chr=="chr14" & modelA_ipDMR$start>=101225830 & modelA_ipDMR$end<=101226250)),]))
modelA_DMR_CpGs<-modelA_DMR_CpGs[,c("probe", "transitioned", "p", "chr","start")]
modelA_DMR_CpGs$Base <- modelA_DMR_CpGs$start+1
modelA_DMR_CpGs <- subset(modelA_DMR_CpGs, select =-c(start))
write.csv(modelA_DMR_CpGs, "Model A/modelA_DMR_CpGs.csv")

##model B
modelB_DMR_CpGs<-(na.omit(modelB_ipDMR[((modelB_ipDMR$chr=="chr5" & modelB_ipDMR$start>=23507130 & modelB_ipDMR$end<=23507400) |(modelB_ipDMR$chr=="chr7" & modelB_ipDMR$start>=38464800 & modelB_ipDMR$end<=38465510)                                          |(modelB_ipDMR$chr=="chr10" & modelB_ipDMR$start>=71323360 & modelB_ipDMR$end<=71324200)|(modelB_ipDMR$chr=="chr14" & modelB_ipDMR$start>=101225830 & modelB_ipDMR$end<=101226250)),]))
modelB_DMR_CpGs<-modelB_DMR_CpGs[,c("probe", "transitioned", "p", "chr","start")]
modelB_DMR_CpGs$Base <- modelB_DMR_CpGs$start+1
modelB_DMR_CpGs <- subset(modelB_DMR_CpGs, select =-c(start))
write.csv(modelB_DMR_CpGs, "Model B/modelB_DMR_CpGs.csv")




```
