---
title: "Power calculations for t-test for Amin's project"
output: html_notebook
---

### Description
Here we calculate the effect sizes for our data, which we can detect with 80% power using t-test for beta-values adjusted using models A and B.
We also check the power to detect differences (using t-test) in a subset of CpGs, which came from the literature.

## Loading libraries
```{r}
library(data.table)
library(pwr)
library(ggplot2)
library(ggpubr)
```

## Setting paths
Change to your paths (folders should exist)

```{r}
data_path = "~/YOUR/PATH/"
results_path = "~/YOUR/PATH/"
images_path = "~/YOUR/PATH/"
```

## Loading the data
```{r}
load(paste(data_path,"adjusted_modelA_sd.RData",sep = ""))
load(paste(data_path,"adjusted_modelB_sd.RData",sep = ""))

adjusted_modelA_cases_sd = adjusted_modelA_case_sd[[1]]
adjusted_modelA_control_sd = adjusted_modelA_control_sd[[1]]

adjusted_modelB_cases_sd = adjusted_modelB_case_sd[[1]]
adjusted_modelB_control_sd = adjusted_modelB_control_sd[[1]]

```

## Calculating pooled SDs for every CpG
```{r}
n_case_A = sum(grouping1=="1") #number of cases
n_control_A = sum(grouping1=="0") #number of controls
pooled_sd_A = sqrt(((n_case_A-1)*adjusted_modelA_case_sd^2+(n_control_A-1)*adjusted_modelA_control_sd^2)/(n_case_A+n_control_A-2))

n_case_B = sum(grouping2=="1") #number of cases
n_control_B = sum(grouping2=="0") #number of controls
pooled_sd_B = sqrt(((n_case_B-1)*adjusted_modelB_case_sd^2+(n_control_B-1)*adjusted_modelB_control_sd^2)/(n_case_B+n_control_B-2))

```


## Creating data for power analysis
```{r}
ds=data.frame()
ds=rbind(ds, c("A",n_case_A,n_control_A,mean(pooled_sd_A[,1]), median(pooled_sd_A[,1])))
ds=rbind(ds, c("B",n_case_B,n_control_B,mean(pooled_sd_B[,1]), median(pooled_sd_B[,1])))
ds[,2:5] = apply(ds[,2:5], MARGIN = 2, as.numeric)
colnames(ds) = c("Model ID", "n, cases", "n, controls", "Pooled SD, mean","Pooled SD, median" )
ds
```

## Power Analysis
```{r}
p = seq(0.01,1,by = 0.01)

# ModelA
dd_a = p
for (i in 1:length(p)) {
  dd_a[i] = pwr.t2n.test(n1 = n_case_A, n2 = n_control_A,power = p[i],sig.level = 0.05/832813)[[3]]
}
plot_ds_a = as.data.frame(cbind(p,dd_a,dd_a*ds$`Pooled SD, mean`[1],dd_a*ds$`Pooled SD, median`[1]))
names(plot_ds_a) = c("power","effect_size_d","effect_size_sd_mean","effect_size_sd_median")

# ModelB
dd_b = p
for (i in 1:length(p)) {
  dd_b[i] = pwr.t2n.test(n1 = n_case_B, n2 = n_control_B,power = p[i],sig.level = 0.05/832813)[[3]]
}
plot_ds_b = as.data.frame(cbind(p,dd_b,dd_b*ds$`Pooled SD, mean`[2],dd_b*ds$`Pooled SD, median`[2]))
names(plot_ds_b) = c("power","effect_size_d","effect_size_sd_mean","effect_size_sd_median")

# Combined dataset for both models
plot_ds = as.data.frame(cbind(c(rep("A",100),rep("B",100)),rbind(plot_ds_a,plot_ds_b)))
names(plot_ds)[1] = "model"

```

## Plotting power vs effect size
# Plot for Model A
```{r}
plot_a = ggplot(plot_ds_a, aes(y = power, x = effect_size_d)) +
  geom_line() + theme_bw() +
  theme_minimal() +
  geom_hline(yintercept = .8,lty=2, color='blue') +
  #geom_point() +
  labs(title="Model A t-test Power Curve for Power vs Effect Size",
       y="Power",
       x="Effect size, Cohen's d")
plot_a
```
# Plot for Model B
```{r}
plot_b = ggplot(plot_ds_b, aes(y = power, x = effect_size_d)) +
  geom_line() + theme_bw() +
  theme_minimal() +
  geom_hline(yintercept = .8,lty=2, color='blue') +
  #geom_point() +
  labs(title="Model B t-test Power Curve for Power vs Effect Size",
       y="Power",
       x="Effect size, Cohen's d")
plot_b
```
# Two models together 
Probably, the individual plot titles should be removed or shortened (in previous plots)
```{r}
ggarrange(plot_a,plot_b,
           ncol = 2, labels = "AUTO")
ggsave(filename = paste(results_path,"power_curve_models_two_panels.png", sep=""),width = 14, height = 7,units = "in")
```
# All lines on the same plot (for Cohen's d)
```{r}
plot_all = ggplot(plot_ds, aes(y = power, x = effect_size_d, group = model)) +
  geom_line(aes(color=model)) + theme_bw() +
  theme_minimal() +
  geom_hline(yintercept = .8,lty=2, color='blue') +
  #geom_point() +
  labs(title="t-test Power Curves for Power vs Effect Size",
       y="Power",
       x="Effect size, Cohen's d")
ggsave(plot = plot_all,filename = paste(results_path,"power_curve_2models_together.png", sep=""))
plot_all

```

# All lines on the same plot (for mean standard deviation)
```{r}
plot_all_mean = ggplot(plot_ds, aes(y = power, x = effect_size_sd_mean, group = model)) +
  geom_line(aes(color=model)) + theme_bw() +
  theme_minimal() +
  geom_hline(yintercept = .8,lty=2, color='blue') +
  #geom_point() +
  labs(title="t-test Power Curves for Power vs Effect Size",
       y="Power",
       x="Effect size, delta beta (mean SD)")
ggsave(plot = plot_all_mean,filename = paste(images_path,"all_models_mean_sd.png",sep=""))

ggsave(plot = plot_all_mean,filename = paste(results_path,"power_curve_2models_together_mean.png",sep=""))

plot_all_mean

```

## Calculating our observed Cohen's d (from data)

## Loading the data
```{r}
load(paste(data_path,"adjusted_modelA_mean.RData",sep = ""))
load(paste(data_path,"adjusted_modelB_mean.RData",sep = ""))

adjusted_modelA_case_mean = adjusted_modelA_case_mean[[1]]
adjusted_modelA_control_mean = adjusted_modelA_control_mean[[1]]

adjusted_modelB_case_mean = adjusted_modelB_case_mean[[1]]
adjusted_modelB_control_mean = adjusted_modelB_control_mean[[1]]
```

## Calculating effect sizes in our data

```{r}
es_data_a = abs(adjusted_modelA_case_mean - adjusted_modelA_control_mean)/pooled_sd_A[[1]]

es_data_b = abs(adjusted_modelB_case_mean - adjusted_modelB_control_mean)/pooled_sd_B[[1]]

es_data = as.data.frame(cbind(es_data_a, es_data_b))

colnames(es_data) = c("a", "b")
head(es_data)
```

```{r}
sig_threshold = 0.05/832813
power_modelA = pwr.t2n.test(n1 = 73, n2 = 203, sig.level = sig_threshold,power = 0.8)
power_modelB = pwr.t2n.test(n1 = 109, n2 = 203, sig.level = sig_threshold,power = 0.8)

power_modelA
power_modelB

```


```{r}
es_hist_a = 
  ggplot(es_data, aes(x=a))+
  geom_histogram(color="darkblue", fill="lightblue",bins = 80)+
  #geom_histogram(aes(y=..density..),color="darkblue", fill="lightblue")+
  #geom_density(alpha=.5, fill = "#34e8eb") +
  labs(x= "Effect size, Cohen's ð‘‘") +
  geom_vline(aes(xintercept=power_modelA$d), color = "red", size = 1)+
  theme_bw()

es_hist_b = 
  ggplot(es_data, aes(x=b))+
  geom_histogram(color="darkblue", fill="lightblue",bins = 80)+
  #geom_histogram(aes(y=..density..),color="darkblue", fill="lightblue")+
  #geom_density(alpha=.5, fill = "#34e8eb") +
  labs(x= "Effect size, Cohen's ð‘‘") +
  geom_vline(aes(xintercept=power_modelB$d), color = "red", size = 1)+
  theme_bw()

es_hist = ggarrange(es_hist_a, es_hist_b,nrow = 2,  labels = c("A","B")) ## Copy-paste to the console if error occur

ggsave(es_hist, filename = paste(results_path,"es_histograms.png", sep = ""), width = 7, height = 7, units = "in", dpi = 300)  ## Copy-paste to the console if error occur
es_hist

```





## Means and SD for the table
# Cohen's d
```{r}
print(paste(round(mean(es_data_a),5)," (", round(sd(es_data_a),5),")",sep=""))
print(paste(round(mean(es_data_b),5)," (", round(sd(es_data_b),5),")",sep=""))
```
# delta beta
```{r}
print(paste(round(mean(abs(adjusted_modelA_case_mean - adjusted_modelA_control_mean)),5)," (", round(sd(abs(adjusted_modelA_case_mean - adjusted_modelA_control_mean)),5),")",sep=""))
print(paste(round(mean(abs(adjusted_modelB_case_mean - adjusted_modelB_control_mean)),5)," (", round(sd(abs(adjusted_modelB_case_mean - adjusted_modelB_control_mean)),5),")",sep=""))
```
# Number of CpGs with Cohen's d above returned by power analysis
```{r}
print(sum(es_data_a>power_modelA$d))
print(sum(es_data_b>power_modelB$d))
```
# Numbers of CpGs with delta betas above returned by power analysis
```{r}
print(sum(abs(adjusted_modelA_case_mean - adjusted_modelA_control_mean)>0.00133))
print(sum(abs(adjusted_modelB_case_mean - adjusted_modelB_control_mean)>0.00117))
```

## Working with CpGs from literature

Loading the data
```{r}
# Loading list of all our CpGs and CpGs from the literature
cpgs_list = (readRDS(paste(data_path,"list_of_CpGs.rds", sep="")))[[1]] # list of all our CpGs - change to your file
cpgs_from_papers = fread(paste(data_path,"cpgs_from_literature_exposures.csv",sep=""), header = TRUE) # CpGs from the literature - change to your file
```


```{r}
cpgs_from_papers$effect_size = as.numeric(cpgs_from_papers$effect_size)

head(cpgs_list)
print("Head of our CpGs list")

head(cpgs_from_papers)
print("Head of CpGs from papers table")

```

List of CpGs from literature missing in our data
```{r}
cpgs_from_papers$cpg[!(cpgs_from_papers$cpg %in% cpgs_list)]
print(paste("In total we are missing ", sum(!(cpgs_from_papers$cpg %in% cpgs_list))," CpGs", sep = ""))
```

```{r}
# calculating delta betas in our data
delta_beta_A = abs(adjusted_modelA_case_mean - adjusted_modelA_control_mean)
delta_beta_B = abs(adjusted_modelB_case_mean - adjusted_modelB_control_mean)

###(!!!)###
# calculating effect sizes in cohen d in our data
es_data_A = delta_beta_A[[1]]/pooled_sd_A[[1]]
es_data_B = delta_beta_B[[1]]/pooled_sd_B[[1]]

# compiling a dataset
es_data = data.frame(cpgs_list,delta_beta_A, delta_beta_B,pooled_sd_A[,1],pooled_sd_B[,1], es_data_A, es_data_B)

colnames(es_data)[1] = "cpg"
colnames(es_data)[4:5] = c("pooled_sd_A", "pooled_sd_B")

es_data = es_data[which(es_data$cpg %in% cpgs_from_papers$cpg),]
es_data = es_data[match(cpgs_from_papers$cpg,es_data$cpg),]

es_data = cbind(cpgs_from_papers,es_data[,-1])

es_data$es_papers_A = abs(es_data$effect_size)/es_data$pooled_sd_A
es_data$es_papers_B = abs(es_data$effect_size)/es_data$pooled_sd_B

es_data$power_data_A = c()
es_data$power_data_B = c()

for (i in 1:599) {
  if (!(is.na(es_data$es_data_A[i]))) {
    es_data$power_data_A[i] = 
      (pwr.t2n.test(n1 = ds$`n, case`[1], n2 = ds$`n, controls`[1], d = es_data$es_data_A[i], sig.level = 0.05/832813))$power
    es_data$power_data_B[i] = 
      (pwr.t2n.test(n1 = ds$`n, case`[2], n2 = ds$`n, controls`[2], d = es_data$es_data_B[i], sig.level = 0.05/832813))$power
  } else {
    es_data$power_data_A[i] = NA
    es_data$power_data_B[i] = NA
    }
}

es_data$power_papers_ANC = c()
es_data$power_papers_BNC = c()

for (i in 1:599) {
  if (!(is.na(es_data$es_papers_A[i])) & !(is.nan(es_data$es_papers_A[i]))) {
    es_data$power_papers_ANC[i] = 
      (pwr.t2n.test(n1 = ds$`n, case`[1], n2 = ds$`n, controls`[1], d = es_data$es_papers_A[i], sig.level = 0.05))$power
    es_data$power_papers_BNC[i] = 
      (pwr.t2n.test(n1 = ds$`n, case`[2], n2 = ds$`n, controls`[2], d = es_data$es_papers_B[i], sig.level = 0.05))$power
  } else {
    es_data$power_papers_ANC[i] = NA
    es_data$power_papers_BNC[i] = NA
    }
}

for (i in 1:599) {
  if (!(is.na(es_data$es_papers_A[i])) & !(is.nan(es_data$es_papers_A[i]))) {
    es_data$power_papers_A413[i] = 
      (pwr.t2n.test(n1 = ds$`n, case`[1], n2 = ds$`n, controls`[1], d = es_data$es_papers_A[i], sig.level = 0.05/413))$power
    es_data$power_papers_B413[i] = 
      (pwr.t2n.test(n1 = ds$`n, case`[2], n2 = ds$`n, controls`[2], d = es_data$es_papers_B[i], sig.level = 0.05/413))$power
  } else {
    es_data$power_papers_A413[i] = NA
    es_data$power_papers_B413[i] = NA
    }
}

fwrite(as.data.frame(es_data), paste(results_path,"calculated_powers_papers.csv", sep=""),col.names = TRUE)
```

