---
title: "R Notebook"
output: html_notebook
---

Step 1: Load libraries

```{r}
library(data.table)
library(ggplot2)
library(xlsx)
library(dplyr)
library(tidyverse)
library(tibble)
```


Step 2: Load CpG files from models
```{r}
## Load Model A
modelA_target <- modelA
modelA_target <- mutate(modelA_target, Significant = "N")  
modelA_target <- dplyr::rename(modelA_target,  "p-value" = "Pr(>|t|)")
modelA_target$Significant[modelA_target$`p-value` <=0.05] <- "Y"
modelA_target$lower <- formatC(modelA_target$lower, format = "e", digits = 2)
modelA_target$upper <- formatC(modelA_target$upper, format = "e", digits = 2)
modelA_target$interaction <- formatC(modelA_target$case, format = "e", digits = 2)
modelA_target$`p-value` <- formatC(modelA_target$`p-value`, format = "e", digits = 2)

## Load Model B
modelB_target <- modelB
modelB_target <- mutate(modelB_target, Significant = "N")  
modelB_target <- dplyr::rename(modelB_target,  "p-value" = "Pr(>|t|)")
modelB_target$Significant[modelB_target$`p-value` <=0.05] <- "Y"
modelB_target$lower <- formatC(modelB_target$lower, format = "e", digits = 2)
modelB_target$upper <- formatC(modelB_target$upper, format = "e", digits = 2)
modelB_target$interaction <- formatC(modelB_target$case, format = "e", digits = 2)
modelB_target$`p-value` <- formatC(modelB_target$`p-value`, format = "e", digits = 2)

```

Step 3: Load CpG manifest and list of exposure CpGs. 

```{r}
##CpG manifest
CpGmanifest <- read.csv("~/data/Round 5 Linear model - interaction variable/infinium-methylationepic-v-1-0-b5-manifest-file.csv")
##Heat stress 
HeatstressCpGs <-read.xlsx("CpG lists/Xu et al, 2021 - Heat - List of CpGs.xlsx", sheetIndex = 2)
HeatstressCpGs365 <- HeatstressCpGs[HeatstressCpGs$Exposure.window=="lag0-365d",]
HeatstressCpGs180 <- HeatstressCpGs[HeatstressCpGs$Exposure.window=="lag0-180d",]

## Organophosphates associated CpGs from Paul et al (2018), PEG study.
OrgPO4_CpGs_paul<-read.xlsx("CpG lists/Paul et al - Organophosphates - 2018.xlsx", sheetIndex = 2)

## Pyrethroid associated CpGs from Furlong et al (2020), PEG study
Pyrethroid_CpGs_furlong <-read.xlsx("CpG lists/Furlong et al - pyrethroid exposure - 2020.xlsx", sheetIndex = 2)

## Multiple pesticide associated CpGs from Hoang et al (2021)
Pesticides_hoang<-read.xlsx("CpG lists/Hoang et al - Multiple pesticides - 2021.xlsx", sheetIndex = 2)

## Arsenic CpGs from Argos et al (2015)
Arsenic_Argos <-  read.xlsx("CpG lists/Argos et al, 2014, Chornic Arsenic exposure EWAS.xlsx", sheetIndex = 2)

## Arsenic CpGs from Ameer et al (2017)
Arsenic_Ameer <- read.xlsx("CpG lists/Arsenic assocaite CpGs, Ameer et al, 2017.xlsx", sheetIndex = 2)

## Arsenic CpGs from Demanelis et al (2019) - Heals and Meta-analysis
Arsenic_Demanelis <- read.xlsx("CpG lists/Demanelis et al, CpGs 2019.xlsx", sheetIndex = 2)

## Arsenic CpGs from Bozack et al (2020) 
Arsenic_Bozack <- read.xlsx("CpG lists/Arsenic associated Cpgs, Bozack et al (Cardenas paper), 2020.xlsx",sheetIndex = 2 )

## Chromium CpGs from Gent et al (2020)
Chromium_Feng <- read.xlsx("CpG lists/Chormium CpGs. Feng et al 2019.xlsx", sheetIndex = 2)
Chromium_Feng <- subset(Chromium_Feng, select = -c(NA., NA..1))

## Cadmium CpGs from Domingo-Relloso et al (2020)
Cadmium_Domingo <- read.xlsx("CpG lists/Domingo-Relloso et al 2020 - Cadmium CpGs.xlsx", sheetIndex = 2)
```

Step 4: Interrogation of heat-stress associated CpGs (Xu et al) 
```{r}
## 365 days of heat
Heat_365_modelA <- left_join(HeatstressCpGs365, modelA_target[, c("rowname","transitioned", "lower","upper","p-value", "Significant")], by = c("CpG" = "rowname"))
Heat_365_modelA <- drop_na(Heat_365_modelA, Significant)
Heat_365_modelA <- arrange(Heat_365_modelA, desc(Significant))
Heat_365_modelA <- mutate(Heat_365_modelA, `Beta congruent and significant` ="N" )
Heat_365_modelA$`Beta congruent and significant`[(Heat_365_modelA$Beta=="+" & Heat_365_modelA$transitioned>0 & Heat_365_modelA$Significant=="Y")|(Heat_365_modelA$Beta=="-" & Heat_365_modelA$transitioned<0 & Heat_365_modelA$Significant=="Y")]  <- "Y"


Heat_365_modelB <- left_join(HeatstressCpGs365,modelB_target[,c("rowname","transitioned","lower", "upper","p-value", "Significant")], by = c("CpG" = "rowname"))
Heat_365_modelB <- drop_na(Heat_365_modelB, Significant)
Heat_365_modelB <- arrange(Heat_365_modelB, desc(Significant))
Heat_365_modelB <- mutate(Heat_365_modelB, `Beta congruent and significant` ="N" )
Heat_365_modelB$`Beta congruent and significant`[(Heat_365_modelB$Beta=="+" & Heat_365_modelB$transitioned>0 & Heat_365_modelB$Significant=="Y")|(Heat_365_modelB$Beta=="-" & Heat_365_modelB$transitioned<0 & Heat_365_modelB$Significant=="Y")]  <- "Y"

## 180 days of heat
Heat_180_modelA <- left_join(HeatstressCpGs180, modelA_target[, c("rowname","transitioned", "lower","upper","p-value", "Significant")], by = c("CpG" = "rowname"))
Heat_180_modelA <- drop_na(Heat_180_modelA, Significant)
Heat_180_modelA <- arrange(Heat_180_modelA, desc(Significant))
Heat_180_modelA <- mutate(Heat_180_modelA, `Beta congruent and significant` ="N" )
Heat_180_modelA$`Beta congruent and significant`[(Heat_180_modelA$Beta=="+" & Heat_180_modelA$transitioned>0 & Heat_180_modelA$Significant=="Y")|(Heat_180_modelA$Beta=="-" & Heat_180_modelA$transitioned<0 & Heat_180_modelA$Significant=="Y")]  <- "Y"

Heat_180_modelB <- left_join(HeatstressCpGs180,modelB_target[,c("rowname","transitioned","lower", "upper","p-value", "Significant")], by = c("CpG" = "rowname"))
Heat_180_modelB <- drop_na(Heat_180_modelB, Significant)
Heat_180_modelB <- arrange(Heat_180_modelB, desc(Significant))
Heat_180_modelB <- mutate(Heat_180_modelB, `Beta congruent and significant` ="N" )
Heat_180_modelB$`Beta congruent and significant`[(Heat_180_modelB$Beta=="+" & Heat_180_modelB$transitioned>0 & Heat_180_modelB$Significant=="Y")|(Heat_180_modelB$Beta=="-" & Heat_180_modelB$transitioned<0 & Heat_180_modelB$Significant=="Y")]  <- "Y"


## Write CSVs
write.csv(Heat_365_modelA, "Model targeted CpGs/heat365_modelA.csv")
write.csv(Heat_365_modelB, "Model targeted CpGs/heat365_modelB.csv")

```

Step 5: Interrogation of pesticide associated CpGs p-values.
```{r}
## Paul et al, Organophosphates. 
OrgPO4_modelA <- left_join(OrgPO4_CpGs_paul, modelA_target[, c("rowname","transitioned","lower", "upper","p-value", "Significant")], by = c("CpG" = "rowname"))
OrgPO4_modelA <- drop_na(OrgPO4_modelA, Significant)
OrgPO4_modelA <- arrange(OrgPO4_modelA, desc(Significant))
OrgPO4_modelA <- mutate(OrgPO4_modelA, `Beta congruent and significant` ="N" )
OrgPO4_modelA$`Beta congruent and significant`[(OrgPO4_modelA$Beta=="+" & OrgPO4_modelA$transitioned>0 & OrgPO4_modelA$Significant=="Y")|(OrgPO4_modelA$Beta=="-" & OrgPO4_modelA$transitioned<0 & OrgPO4_modelA$Significant=="Y")]  <- "Y"


OrgPO4_modelB <- left_join(OrgPO4_CpGs_paul, modelB_target[, c("rowname","transitioned","lower", "upper","p-value", "Significant")], by = c("CpG" = "rowname"))
OrgPO4_modelB <- drop_na(OrgPO4_modelB, Significant)
OrgPO4_modelB <- arrange(OrgPO4_modelB, desc(Significant))
OrgPO4_modelB <- mutate(OrgPO4_modelB, `Beta congruent and significant` ="N" )
OrgPO4_modelB$`Beta congruent and significant`[(OrgPO4_modelB$Beta=="+" & OrgPO4_modelB$transitioned>0 & OrgPO4_modelB$Significant=="Y")|(OrgPO4_modelB$Beta=="-" & OrgPO4_modelB$transitioned<0 & OrgPO4_modelB$Significant=="Y")]  <- "Y" 


## Furlong et al, Pyrethroid. 
Pyrethroid_modelA <- left_join(Pyrethroid_CpGs_furlong , modelA_target[, c("rowname","transitioned","lower","upper","p-value", "Significant")], by = c("CpG" = "rowname"))
Pyrethroid_modelA <- drop_na(Pyrethroid_modelA, Significant)
Pyrethroid_modelA <- arrange(Pyrethroid_modelA, desc(Significant))
Pyrethroid_modelA <- mutate(Pyrethroid_modelA, `Beta congruent and significant` ="N" )
Pyrethroid_modelA$`Beta congruent and significant`[(Pyrethroid_modelA$Beta=="+" & Pyrethroid_modelA$transitioned>0 & Pyrethroid_modelA$Significant=="Y")|(Pyrethroid_modelA$Beta=="-" & Pyrethroid_modelA$transitioned<0 & Pyrethroid_modelA$Significant=="Y")]  <- "Y"


Pyrethroid_modelB <- left_join(Pyrethroid_CpGs_furlong , modelB_target[, c("rowname","transitioned","lower","upper","p-value", "Significant")], by = c("CpG" = "rowname"))
Pyrethroid_modelB <- drop_na(Pyrethroid_modelB, Significant)
Pyrethroid_modelB <- arrange(Pyrethroid_modelB, desc(Significant))
Pyrethroid_modelB <- mutate(Pyrethroid_modelB, `Beta congruent and significant` ="N" )
Pyrethroid_modelB$`Beta congruent and significant`[(Pyrethroid_modelB$Beta=="+" & Pyrethroid_modelB$transitioned>0 & Pyrethroid_modelB$Significant=="Y")|(Pyrethroid_modelB$Beta=="-" & Pyrethroid_modelB$transitioned<0 & Pyrethroid_modelB$Significant=="Y")]  <- "Y" 



## Hoang et al, multiple pesticides
Mixed_pesticides_modelA <- left_join(Pesticides_hoang, modelA_target[, c("rowname","transitioned","lower","upper","p-value", "Significant")], by = c("CpG" = "rowname"))
Mixed_pesticides_modelA <- drop_na(Mixed_pesticides_modelA, Significant)
mixed_pesticides_modelA <- arrange(Mixed_pesticides_modelA, desc(Significant))
Mixed_pesticides_modelA <- mutate(Mixed_pesticides_modelA, `Beta congruent and significant` ="N" )
Mixed_pesticides_modelA$`Beta congruent and significant`[(Mixed_pesticides_modelA$Beta=="+" & Mixed_pesticides_modelA$transitioned>0 & Mixed_pesticides_modelA$Significant=="Y")|(Mixed_pesticides_modelA$Beta=="-" & Mixed_pesticides_modelA$transitioned<0 & Mixed_pesticides_modelA$Significant=="Y")]  <- "Y"


Mixed_pesticides_modelB <- left_join(Pesticides_hoang, modelB_target[, c("rowname","transitioned","lower","upper","p-value", "Significant")], by = c("CpG" = "rowname"))
Mixed_pesticides_modelB <- drop_na(Mixed_pesticides_modelB, Significant)
Mixed_pesticides_modelB <- arrange(Mixed_pesticides_modelB, desc(Significant))
Mixed_pesticides_modelB <- mutate(Mixed_pesticides_modelB, `Beta congruent and significant` ="N" )
Mixed_pesticides_modelB$`Beta congruent and significant`[(Mixed_pesticides_modelB$Beta=="+" & Mixed_pesticides_modelB$transitioned>0 & Mixed_pesticides_modelB$Significant=="Y")|(Mixed_pesticides_modelB$Beta=="-" & Mixed_pesticides_modelB$transitioned<0 & Mixed_pesticides_modelB$Significant=="Y")]   <- "Y"


## Write CSVs
write.csv(OrgPO4_modelA, "Model targeted CpGs/OrgPO4_modelA.csv")
write.csv(OrgPO4_modelB, "Model targeted CpGs/OrgPO4_modelB.csv")
write.csv(Pyrethroid_modelA, "Model targeted CpGs/Pyrethroid_modelA.csv")
write.csv(Pyrethroid_modelB, "Model targeted CpGs/Pyrethroid_modelB.csv")
write.csv(Mixed_pesticides_modelA, "Model targeted CpGs/mixed_pesticides_modelA.csv")
write.csv(Mixed_pesticides_modelB, "Model targeted CpGs/mixed_pesticides_modelB.csv")
```

Step 6: Interrogation of Arsenic associated CpGs. 
```{r}
## Argos et al (2015)
Arsenic_Argos_modelA <- left_join(Arsenic_Argos, modelA_target[, c("rowname","transitioned","lower","upper","p-value", "Significant")], by = c("CpG" = "rowname"))
Arsenic_Argos_modelA <- drop_na(Arsenic_Argos_modelA, Significant)
Arsenic_Argos_modelA <- arrange(Arsenic_Argos_modelA, desc(Significant))
Arsenic_Argos_modelA <- mutate(Arsenic_Argos_modelA, `Beta congruent and significant` ="N" )
Arsenic_Argos_modelA$`Beta congruent and significant`[(Arsenic_Argos_modelA$Beta=="+" & Arsenic_Argos_modelA$transitioned>0 & Arsenic_Argos_modelA$Significant=="Y")|(Arsenic_Argos_modelA$Beta=="-" & Arsenic_Argos_modelA$transitioned<0 & Arsenic_Argos_modelA$Significant=="Y")]  <- "Y"


Arsenic_Argos_modelB <- left_join(Arsenic_Argos, modelB_target[, c("rowname","transitioned","lower","upper","p-value", "Significant")], by = c("CpG" = "rowname"))
Arsenic_Argos_modelB <- drop_na(Arsenic_Argos_modelB, Significant)
Arsenic_Argos_modelB <- arrange(Arsenic_Argos_modelB, desc(Significant))
Arsenic_Argos_modelB <- mutate(Arsenic_Argos_modelB, `Beta congruent and significant` ="N" )
Arsenic_Argos_modelB$`Beta congruent and significant`[(Arsenic_Argos_modelB$Beta=="+" & Arsenic_Argos_modelB$transitioned>0 & Arsenic_Argos_modelB$Significant=="Y")|(Arsenic_Argos_modelB$Beta=="-" & Arsenic_Argos_modelB$transitioned<0 & Arsenic_Argos_modelB$Significant=="Y")]  <- "Y" 


## Ameer et al (2017)
Arsenic_Ameer_modelA <- left_join(Arsenic_Ameer, modelA_target[, c("rowname","transitioned","lower","upper","p-value", "Significant")], by = c("CpG" = "rowname"))
Arsenic_Ameer_modelA <- drop_na(Arsenic_Ameer_modelA, Significant)
Arsenic_Ameer_modelA <- arrange(Arsenic_Ameer_modelA, desc(Significant))
Arsenic_Ameer_modelA <- mutate(Arsenic_Ameer_modelA, `Beta congruent and significant` ="N" )
Arsenic_Ameer_modelA$`Beta congruent and significant`[(Arsenic_Ameer_modelA$Beta=="+" & Arsenic_Ameer_modelA$transitioned>0 & Arsenic_Ameer_modelA$Significant=="Y")|(Arsenic_Ameer_modelA$Beta=="-" & Arsenic_Ameer_modelA$transitioned<0 & Arsenic_Ameer_modelA$Significant=="Y")]  <- "Y"


Arsenic_Ameer_modelB <- left_join(Arsenic_Ameer, modelB_target[, c("rowname","transitioned","lower","upper","p-value", "Significant")], by = c("CpG" = "rowname"))
Arsenic_Ameer_modelB <- drop_na(Arsenic_Ameer_modelB, Significant) 
Arsenic_Ameer_modelB <- arrange(Arsenic_Ameer_modelB, desc(Significant))
Arsenic_Ameer_modelB <- mutate(Arsenic_Ameer_modelB, `Beta congruent and significant` ="N" )
Arsenic_Ameer_modelB$`Beta congruent and significant`[(Arsenic_Ameer_modelB$Beta=="+" & Arsenic_Ameer_modelB$transitioned>0 & Arsenic_Ameer_modelB$Significant=="Y")|(Arsenic_Ameer_modelB$Beta=="-" & Arsenic_Ameer_modelB$transitioned<0 & Arsenic_Ameer_modelB$Significant=="Y")]  <- "Y" 


## Demanelis et al (2019) 
Arsenic_Demanelis_modelA <- left_join(Arsenic_Demanelis, modelA_target[, c("rowname","transitioned","lower","upper","p-value", "Significant")], by = c("CpG" = "rowname"))
Arsenic_Demanelis_modelA <- drop_na(Arsenic_Demanelis_modelA, Significant)
Arsenic_Demanelis_modelA <- arrange(Arsenic_Demanelis_modelA, desc(Significant))
Arsenic_Demanelis_modelA <- mutate(Arsenic_Demanelis_modelA, `Beta congruent and significant` ="N" )
Arsenic_Demanelis_modelA$`Beta congruent and significant`[(Arsenic_Demanelis_modelA$Beta=="+" & Arsenic_Demanelis_modelA$transitioned>0 & Arsenic_Demanelis_modelA$Significant=="Y")|(Arsenic_Demanelis_modelA$Beta=="-" & Arsenic_Demanelis_modelA$transitioned<0 & Arsenic_Demanelis_modelA$Significant=="Y")]  <- "Y"


Arsenic_Demanelis_modelB <- left_join(Arsenic_Demanelis, modelB_target[, c("rowname","transitioned","lower","upper","p-value", "Significant")], by = c("CpG" = "rowname"))
Arsenic_Demanelis_modelB <- drop_na(Arsenic_Demanelis_modelB, Significant)
Arsenic_Demanelis_modelB <- arrange(Arsenic_Demanelis_modelB, desc(Significant))
Arsenic_Demanelis_modelB <- mutate(Arsenic_Demanelis_modelB, `Beta congruent and significant` ="N" )
Arsenic_Demanelis_modelB$`Beta congruent and significant`[(Arsenic_Demanelis_modelB$Beta=="+" & Arsenic_Demanelis_modelB$transitioned>0 & Arsenic_Demanelis_modelB$Significant=="Y")|(Arsenic_Demanelis_modelB$Beta=="-" & Arsenic_Demanelis_modelB$transitioned<0 & Arsenic_Demanelis_modelB$Significant=="Y")]  <- "Y" 


## Bozack et al (2020)
Arsenic_Bozack_modelA <- left_join(Arsenic_Bozack, modelA_target[, c("rowname","transitioned","lower","upper","p-value", "Significant")], by = c("CpG" = "rowname"))
Arsenic_Bozack_modelA <- drop_na(Arsenic_Bozack_modelA, Significant)
Arsenic_Bozack_modelA <- arrange(Arsenic_Bozack_modelA, desc(Significant))
Arsenic_Bozack_modelA <- mutate(Arsenic_Bozack_modelA, `Beta congruent and significant` ="N" )
Arsenic_Bozack_modelA$`Beta congruent and significant`[(Arsenic_Bozack_modelA$Beta=="+" & Arsenic_Bozack_modelA$transitioned>0 & Arsenic_Bozack_modelA$Significant=="Y")|(Arsenic_Bozack_modelA$Beta=="-" & Arsenic_Bozack_modelA$transitioned<0 & Arsenic_Bozack_modelA$Significant=="Y")]  <- "Y"


Arsenic_Bozack_modelB <- left_join(Arsenic_Bozack, modelB_target[, c("rowname","transitioned","lower","upper","p-value", "Significant")], by = c("CpG" = "rowname"))
Arsenic_Bozack_modelB <- drop_na(Arsenic_Bozack_modelB, Significant)
Arsenic_Bozack_modelB <- arrange(Arsenic_Bozack_modelB, desc(Significant))
Arsenic_Bozack_modelB <- mutate(Arsenic_Bozack_modelB, `Beta congruent and significant` ="N" )
Arsenic_Bozack_modelB$`Beta congruent and significant`[(Arsenic_Bozack_modelB$Beta=="+" & Arsenic_Bozack_modelB$transitioned>0 & Arsenic_Bozack_modelB$Significant=="Y")|(Arsenic_Bozack_modelB$Beta=="-" & Arsenic_Bozack_modelB$transitioned<0 & Arsenic_Bozack_modelB$Significant=="Y")]  <- "Y" 


## Write CSVs
write.csv(Arsenic_Argos_modelA, "Model targeted CpGs/Arsenic_Argos_modelA.csv")
write.csv(Arsenic_Argos_modelB, "Model targeted CpGs/Arsenic_Argos_modelB.csv")
write.csv(Arsenic_Ameer_modelA, "Model targeted CpGs/Arsenic_Ameer_modelA.csv")
write.csv(Arsenic_Ameer_modelB, "Model targeted CpGs/Arsenic_Ameer_modelA.csv")
write.csv(Arsenic_Demanelis_modelA, "Model targeted CpGs/Arsenic_Demanelis_modelA.csv")
write.csv(Arsenic_Demanelis_modelB, "Model targeted CpGs/Arsenic_Demanelis_modelB.csv")
write.csv(Arsenic_Bozack_modelA, "Model targeted CpGs/Arsenic_Bozack_modelA.csv")
write.csv(Arsenic_Bozack_modelB, "Model targeted CpGs/Arsenic_Bozack_modelB.csv")

```

Step 7: Interrogation of other metals associated CpGs. 
```{r}
## Chromium, Feng et al (2020)
Chromium_Feng_modelA <- left_join(Chromium_Feng, modelA_target[, c("rowname","transitioned","lower","upper","p-value", "Significant")], by = c("CpG" = "rowname"))
Chromium_Feng_modelA <- drop_na(Chromium_Feng_modelA, Significant)
Chromium_Feng_modelA <- arrange(Chromium_Feng_modelA, desc(Significant))
Chromium_Feng_modelA <- mutate(Chromium_Feng_modelA, `Beta congruent and significant` ="N" )
Chromium_Feng_modelA$`Beta congruent and significant`[(Chromium_Feng_modelA$Beta=="+" & Chromium_Feng_modelA$transitioned>0 & Chromium_Feng_modelA$Significant=="Y")|(Chromium_Feng_modelA$Beta=="-" & Chromium_Feng_modelA$transitioned<0 & Chromium_Feng_modelA$Significant=="Y")]  <- "Y"


Chromium_Feng_modelB <- left_join(Chromium_Feng, modelB_target[, c("rowname","transitioned","lower","upper","p-value", "Significant")], by = c("CpG" = "rowname"))
Chromium_Feng_modelB <- drop_na(Chromium_Feng_modelB, Significant)
Chromium_Feng_modelB <- arrange(Chromium_Feng_modelB, desc(Significant))
Chromium_Feng_modelB <- mutate(Chromium_Feng_modelB, `Beta congruent and significant` ="N" )
Chromium_Feng_modelB$`Beta congruent and significant`[(Chromium_Feng_modelB$Beta=="+" & Chromium_Feng_modelB$transitioned>0 & Chromium_Feng_modelB$Significant=="Y")|(Chromium_Feng_modelB$Beta=="-" & Chromium_Feng_modelB$transitioned<0 & Chromium_Feng_modelB$Significant=="Y")]  <- "Y" 



## Cadmium, 
Cadmium_Domingo_modelA <- left_join(Cadmium_Domingo, modelA_target[, c("rowname","transitioned","lower","upper","p-value", "Significant")], by = c("CpG" = "rowname"))
Cadmium_Domingo_modelA <- drop_na(Cadmium_Domingo_modelA, Significant)
Cadmium_Domingo_modelA <- arrange(Cadmium_Domingo_modelA, desc(Significant))
Cadmium_Domingo_modelA <- mutate(Cadmium_Domingo_modelA, `Beta congruent and significant` ="N" )
Cadmium_Domingo_modelA$`Beta congruent and significant`[(Cadmium_Domingo_modelA$Beta=="+" & Cadmium_Domingo_modelA$transitioned>0 & Cadmium_Domingo_modelA$Significant=="Y")|(Cadmium_Domingo_modelA$Beta=="-" & Cadmium_Domingo_modelA$transitioned<0 & Cadmium_Domingo_modelA$Significant=="Y")]  <- "Y"


Cadmium_Domingo_modelB <- left_join(Cadmium_Domingo, modelB_target[, c("rowname","transitioned","lower","upper", "p-value", "Significant")], by = c("CpG" = "rowname"))
Cadmium_Domingo_modelB <- drop_na(Cadmium_Domingo_modelB, Significant)
Cadmium_Domingo_modelB <- arrange(Cadmium_Domingo_modelB, desc(Significant))
Cadmium_Domingo_modelB <- mutate(Cadmium_Domingo_modelB, `Beta congruent and significant` ="N" )
Cadmium_Domingo_modelB$`Beta congruent and significant`[(Cadmium_Domingo_modelB$Beta=="+" & Cadmium_Domingo_modelB$transitioned>0 & Cadmium_Domingo_modelB$Significant=="Y")|(Cadmium_Domingo_modelB$Beta=="-" & Cadmium_Domingo_modelB$transitioned<0 & Cadmium_Domingo_modelB$Significant=="Y")]  <- "Y" 



## Write CSVs
write.csv(Chromium_Feng_modelA, "Model targeted CpGs/Chromium_Feng_modelA.csv")
write.csv(Chromium_Feng_modelB, "Model targeted CpGs/Chromium_Feng_modelB.csv")
write.csv(Cadmium_Domingo_modelA, "Model targeted CpGs/Cadmium_Domingo_modelA.csv")
write.csv(Cadmium_Domingo_modelB, "Model targeted CpGs/Cadmium_Domingo_modelB.csv")

```

Step 8: Write Excel files of all CpGs per model
```{r}
## Model 1
openxlsx::write.xlsx(list( "Sheet1" = Heat_365_modelA, "Sheet2" = Arsenic_Argos_modelA , "Sheet3" = Arsenic_Ameer_modelA, "Sheet4" = Arsenic_Demanelis_modelA, "Sheet5" = Arsenic_Bozack_modelA, "Sheet6" = Cadmium_Domingo_modelA, "Sheet7" =Chromium_Feng_modelA,  "Sheet8" = OrgPO4_modelA ,"Sheet9" = Pyrethroid_modelA, "Sheet10" = Mixed_pesticides_modelA),file = "Model targeted CpGs/ModelA_CpG_dataset.xlsx", rowNames=FALSE, append=TRUE )

## Model 2
openxlsx::write.xlsx(list( "Sheet1" = Heat_365_modelB, "Sheet2" = Arsenic_Argos_modelB , "Sheet3" = Arsenic_Ameer_modelB, "Sheet4" = Arsenic_Demanelis_modelB, "Sheet5" = Arsenic_Bozack_modelB, "Sheet6" = Cadmium_Domingo_modelB, "Sheet7" =Chromium_Feng_modelB,  "Sheet8" = OrgPO4_modelB ,"Sheet9" = Pyrethroid_modelB, "Sheet10" = Mixed_pesticides_modelB),file = "Model targeted CpGs/ModelB_CpG_dataset.xlsx", rowNames=FALSE, append=TRUE )

```

